Determining the existance and direction of causality in the face of measurement error

### 2016-05-09 docs/manuscript.Rmd
finished new draft

tab_nums(
	"", 
	caption="", 
	display=FALSE)

@@

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes)."
)

@@

@@

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible. We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. We argue that because measurement error is ubiquitous in phenotypic data, mediation-based causal inference should be treated with caution.

@@

The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene expression all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, commonly used cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding.

An established statistical design that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant cross sectional data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels. Two statistical approaches to exploiting the properties of genetic instruments are commonly used: mediation and Mendelian randomisation (MR)  [@DaveySmith2004; @DaveySmithHemani2014].

Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008]. A mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable. If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.

This simplistic scenario, however, is seldom borne out in reality. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator. A proposed solution to this is to use a genetic instrument as the independent variable. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

MR employs the same data under a completely different model. Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome. Here the problems of confounding or reverse causation are avoided by testing for association between the outcome and the SNP. The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables.

Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure. This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

MR does however have some important advantages over mediation based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR relax this assumption, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes. Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence. We argue that this is a serious limitation.

Measurement (or observational) error is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference. Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially.

@@

where $y - \hat{y} = y - \hat{alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$. The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.

@@

where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$.

@@

Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common. An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different. The method and R function are provided in Appendix 3.

@@

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and $*_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $*_{my}$ are parameters for non-differential measurement error in the outcome $y$.

All $\alpha$ values were set to 0, and $\beta$ values set to 1. Values of $\epsilon_*$ were generated such that

@@

@@

@@

In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis. Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

@@

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012] network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.

There are a number of limitations to the analysis that we performed. We focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. To give an example in which this might arise consider the case that C-reactive protein (CRP) is being tested for causal association with some outcome, and the FTO variant is used as an instrument. Though FTO is robustly associated with CRP, it is through the upstream influence of BMI [@Timpson2011]. Thus, if BMI has a direct influence on the outcome, the conditions of MR will be violated and the Steiger test proposed here will be unreliable.

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is very susceptible to measurement error. With the presence of measurement error near ubiquitous in most cross sectional studies, and our ability to measure it limited, we argue that it should be given serious consideration in designing statistical analyses that attempt to make causal inference, and that any putitive results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

@@

@@

$$
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_n + \beta_n y + \epsilon_n
\end{aligned}
$$

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from bias in $\alpha_n$ and $\beta_n$ and imprecision from $\epsilon_n. Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$. Under the CIT, this is expected to be $cov(g, y_O - \hat{y}_O) = 0$ when $x$ causes $y$, where $\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O$ is the predicted value of $y_O$ using the measured value of $x_O$.

@@

$$
\begin{aligned}
cov(g, y_O) & = cov(g, \beta_n y) \\
            & = cov(g, \beta_n \beta_x x) \\
            & = cov(g, \beta_n \beta_x \beta_g g) \\
            & = \beta_n \beta_x \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                  & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
                  & = \frac{cov(\beta_n y, \beta_m x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m \beta_n cov(y, x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m \beta_n \beta_x var(x)} {\beta_m^2 var(x) + var(\epsilon_m)}
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = \frac{\beta_n \beta_x \beta_g var(g) \beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \\
                  & = \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, y_O - \hat{y}_O) & = \beta_n \beta_x \beta_g var(g) - \frac{\beta_m^2 var(x)} {\beta_m^2 var(x) + var(\epsilon_m)} \times \beta_n \beta_x \beta_g var(g)
\end{aligned}
$$

>>>>

tab_nums(
	"shakhtab", 
	caption="Summary of MR analysis on gene expression and DNA methylation associations. 'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the Steiger test. P(+ve effect) denotes the proportion of associations that were positive. $cor(\\rho_{MR}, \\rho_{P})$ is the correlation of the MR estimate obtained in this analysis and the Pearson correlation between reported in Shakhbazov et al 2016.", 
	display=FALSE)

@@

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared with for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
	display=FALSE
)

@@

```{r shakhbazov }

@@

With our ability to characterize the human phenome ever improving it is becoming increasingly common to use statistical tests to infer the causal relationships between correlated variables. A simple and commonly used method for inferring if an exposure is on the causal pathway to an outcome is through using genetic instruments in mediation analysis, where the effect of an instrument on the outcome is tested before and after adjusting for the exposure. We show that in the face of measurement error this can lead to erroneous results, and that increasing sample size, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. Commonly used tools that are predicated on this methodology such as the causal inference test (CIT) are shown to be susceptible. We demonstrate that Mendelian randomization (MR) is a method for causal inference that is robust to measurement error, and that a simple extension to MR can provide a method to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. Using this method we demonstrate that for correlations between DNA methylation and gene expression levels, DNA methylation is likely to be the causal factor. We show that using MR to infer causal directions is straightforward, but because measurement error is ubiquitous in observational data, mediation-based causal inference in particular should be treated with caution.

@@

The last decade has seem a rapid acceleration in technological advances that enable us to measure the human phenome in ever increasing detail, furnishing biologists with the tantalising opportunity to map the molecular basis of complex traits. This does, however, require overcoming the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, identifying metabolic perturbations that lead to disease [@Kettunen2012], or understanding the influence of epigenomic influences on gene expression [@Bird2002] all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, commonly used observational study data are notoriously susceptible to pitfalls in making causal inference, with simple regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@DaveySmith2004].

An established statistical design that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the causal nature of the association by randomly splitting the sample into two groups, subjecting one group to the exposure and leaving the other as a control. The association between the exposure and the outcome in this setting provides a robust estimate of the causal relationship. This is the theoretical basis behind randomised control trials, but in practice randomisation is often impossible to implement in an experimental context due to cost, scale or inability to manipulate the exposure. The principle, however, can be employed in extant observational data through the use of genetic variants associated with the exposure (instruments), where the inheritance of an allele serves as a random lifetime allocation of differential exposure levels [@DaveySmith2004]. Two statistical approaches to exploiting the properties of genetic instruments are widely used: mediation and Mendelian randomisation (MR).

Mediation is a commonly used statistical method for making causal inference in observational data, and it exists in many forms, from regression based systems  [@Baron1986] to structural equation modeling (SEM) [@Aten2008]. In its simplest form mediation analysis seeks to explain the mechanism that underlies an observed relationship between an independent and a dependent variable by introducing a third, 'mediator' variable. If the association between the independent and dependent variables disappears when conditioning on the mediator variable then this provides evidence that the independent variable influences the dependent variable through the mediator.

This simple form, however, is prone to being unreliable. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator [@Fewell2007]. Introducing pseudo-randomisation by using a genetic instrument as the independent variable can be used to improve on these issues. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods such as genetical genomics [@Schadt2005], the regression-based causal inference test (CIT) [@Millstein2009; @Millstein2016], an SEM implementation in the NEO software [@Aten2008], and various other methods including Bayesian approaches [@Waszak2015]. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015; @Yuan2014; @Tang2014; @Hong2015].

MR employs the same data under but uses a different approach in how the instrumental variable is employed. Here an instrumental variable (typically a SNP) that has a robost, direct association with the exposure is used as a surrogate for the exposure in an association test with the outcome, avoiding the problems of confounding or reverse causation by testing for association between the outcome and the SNP. The causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation and MR models have properties that confer some advantages and some disadvantages. The main advantage of mediation over MR is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence.

Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure. This requirement can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. Another situation is where the causal directions between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016]?

MR does however have some important advantages over mediation based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, using mediation requires that the exposure, outcome and instrument variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes. Third, and though this is often underappreciated in fields outside classical epidemiology it is of crucial importance, MR is substantially more robust to there being measurement error in the exposure and outcome variables. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error [@Ashenfelter1994], whereas it has been noted that both basic mediation [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence. We argue that this is a serious limitation.

Measurement (or observational) error is the difference between the measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more low-tech data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiologyical literature extensively [@Fewell2007; @LeCessie2012; @Blakely2013].

Given the near-ubiquitous presence of measurement error in biological data it is vital to understand its impact on the tools we use for causal inference. Here we show using theory and simulations how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We also present an extension to MR that allows researchers to ascertain the causal direction even when the biology of the instrument is not fully understood. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially, and we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for the $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$. The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.

@@

where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].

@@

Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting. Two-sample MR refers to the case where the SNP-exposure association and SNP-outcome association are calculated in different samples (or from publicly available summary statistics). Here the Steiger test of two independent correlations can be applied where.

$$
Z = \frac{Z_{gx} - Z_{gy}} { \sqrt{ 1 / (N_{1} - 3) + 1 / (N_{2} - 3) } }
$$

An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different. Steiger test statistics were calculated using the *r.test* function in the R package *R/psych* [@Revelle2015].

@@

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.

All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distribted values of $\epsilon_*$ were generated such that

@@

## Two sample MR

Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels. Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations. The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.

@@

### MR analysis of gene expression and DNA methylation

We used the Steiger test to test if DNA methylation levels are more likely to causally influence gene expression levels or vice versa. To do this we obtained a list of `r nrow(qtl)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was always in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], and that there was a moderate correlation between the magnitude of the causal correlation and the observational Pearson correlation (`r ctab("shakhtab")`).

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`)), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

In the epidemiological literature issues of measurement error in mediation is relatively well explored [@Fewell2007]. Our analysis extends this to related  methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

@@

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a], network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.

In our simulations we focused on the simple case of a single instrument in a single sample setting, and assumed that pleiotropy (the influence of the instrument on the outcome through a mechanism other than the exposure) was not present. Recent method developments in MR [@Bowden2015] have focused on accounting for the issues that horizontal pleiotropy can introduce, but how they perform in the presence of measurement error remains to be explored. An important advantage that MR confers over most mediation based analysis is that it can be performed in two samples, which can considerably improve power and expand the scope of analysis. However, whether there is a substantive difference in two sample MR versus one sample MR in how measurement error has an effect is not yet fully understood. We have also assumed no measurement error in the genetic instrument, which is not unreasonable given the strict QC protocols that ensure high quality genotype data is available to most studies. We have restricted the scope to only exploring non-differential measurement error and avoided the complications incurred if measurement error in the exposure and outcome is correlated, however our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

Our analysis of DNA methylation and gene expression summary data demonstates the simplicity of inferring causal existence, direction and magnitude in the two sample MR framework. We note that verifying these results with regards to the extent to which measurement error has influenced them empirically is a challenging task and beyond the scope of this work.

The overarching result from our simulations is that, regardless of the method used, inferring causal direction using an instrument of unknown biology is highly sensitive to measurement error. With the presence of measurement error near ubiquitous in most observational data, and our ability to measure it limited, we argue that it needs to be central to any consideration of approaches which are used in attempt to strengthen causal inference, and any putative results should be accompanied with appropriate sensitivity analysis that assesses their robustness under varying levels of measurement error.

## Tables

```{r shakhtab }

`r tab_nums("shakhtab")`

@@

```{r causality_exists_tpr, fig.width=10, fig.height=10 }

`r fig_nums("causality_exists_tpr")`

@@

$$
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_{my} + \beta_{my} y + \epsilon_{my}
\end{aligned}
$$

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where measurement error is incurred from linear transformation in $\alpha_{mx}$ and $\beta_{mx}$ and imprecision from $\epsilon_{mx}$, and $y_O$ is the measured quantity of $y$, where measurement error is incurred from linear transformation in $\alpha_{my}$ and $\beta_{my}$ and imprecision from $\epsilon_{my}$. Our objective is to estimate the expected magnitude of association between $g$ and $y$ after conditioning on $x$. Under the CIT, this is expected to be $cov(g, y_O - \hat{y}_O) = 0$ when $x$ causes $y$, where $\hat{y}_O = \hat{a}_{x_O} + \hat{\beta}_{x_O} x_O$ is the predicted value of $y_O$ using the measured value of $x_O$.

@@

$$
\begin{aligned}
cov(g, y_O) & = cov(g, \beta_{my} y) \\
            & = cov(g, \beta_{my} \beta_x x) \\
            & = cov(g, \beta_{my} \beta_x \beta_g g) \\
            & = \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} x) \\
                  & = cov(g, \hat{\beta}_{x_O} \beta_{mx} \beta_g g) \\
                  & = \hat{\beta}_{x_O} \beta_{mx} \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
\hat{\beta}_{x_O} & = \frac{cov(y_O, x_O)} {var(x_O)} \\
                  & = \frac{cov(\beta_{my} y, \beta_{mx} x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} cov(y, x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx} \beta_{my} \beta_x var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})}
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, \hat{y}_O) & = \frac{\beta_{my} \beta_x \beta_g var(g) \beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \\
                  & = \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_{mx})} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$

@@

$$
\begin{aligned}
cov(g, y_O - \hat{y}_O) & = \beta_{my} \beta_x \beta_g var(g) - \frac{\beta_{mx}^2 var(x)} {\beta_{mx}^2 var(x) + var(\epsilon_m)} \times \beta_{my} \beta_x \beta_g var(g)
\end{aligned}
$$
